Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2007-07-04 SCP2008/paper/metamonads.tex
updating the text with the new version of the code

We have several Rank aspects:
\begin{code2}
module Rank = struct
  include TrackRank
  let fin () = perform
      r <-- mo_lookup ip;
      ret (liftGet r)
end

module NoRank = struct
  include TrackRank
  let fin () = failwith "Rank is needed but is not computed"
end
\end{code2}

@@

As is apparent from the output choices, several different quantities
\emph{may} need to be tracked in a particular LU implementation.  We
therefore need to be able to conditionally generate variables
representing the tracking state, and weave in corresponding tracking
code. We may need to (independently) keep track of the rank, the
determinant and the permutation list.  The tracking state variables
then become part of the \emph{state} that is tracked by our monad.  To
have all this choice when needed, and yet have our code be modular and
composable as well as ensuring that the generated code does not
contain any abstraction artifacts, it is important to make this state
modular.  For example,
\begin{code}
module type DETERMINANT = sig
  type indet  type outdet  type 'a lstate
  type tdet = outdet ref   
  val decl : unit -> 
    (unit, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
  val upd_sign : unit -> 
    (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
  ...
end
\end{code}
\noindent  to track determinant we should be able to generate code
for: defining variables used for tracking (|decl|),
updating the sign or the absolute
value of the determinant, converting the tracking state
to the final determinant value of the type |outdet|. LU of a
floating-point matrix with no determinant tracking uses the
instantiation of |DETERMINANT| where |outdet| is |unit| and all the
functions of that module generate no code. For integer matrices, we
have to track some aspects of the determinant, even if we don't output
it. The determinant tracking aspect is complex because tracking
variables, if any, are to be declared at the beginning of LU; the sign
of the determinant has to be updated on each row or column
permutation; the value of the determinant should be updated per each
pivoting. We use |lstate| to pass the tracking state, e.g., a piece of
code for the value of the type |Dom.v ref|, among
various determinant-tracking functions. The |lstate| is a part of the
overall monadic state. Other aspects, e.g., rank tracking, may use the
monadic state for passing of rank tracking variables. To be able to
compose determinant and rank tracking functors -- each of which may
(or may not) use the monadic state for passing its own data -- we make
extensive use of open records (a list of polymorphic variants
appeared to be the easiest way to implement such a union, in a purely
functional way). This lets us freely compose determinant-tracking,
rank-tracking, and other aspects.

@@

\begin{figure}
\begin{code2}
module GenGE(F : FEATURES) = struct
    module O = F.Output(F)

@@

   let gen input = perform
          (mat, r, c, rmar) <-- init input;
          seqM 
            (forward_elim (mat, r, c, rmar))
            (O.make_result mat)
end
\end{code2}
\end{figure}

@@

In addition to the ``regular'' type sharing constraints shown in the
|Gen| functor, there are also ``semantic'' sharing constraints, shown
in the following structure of the |UPDATE| signature:
\vspace*{-5pt}\begin{code}
module DivisionUpdate
  (Dom:DOMAIN with type kind = domain_is_field)
  (C:CONTAINER2D)
  (Det:DETERMINANT with type indet=Dom.v) = struct ... end
\end{code}
\vspace*{-4pt} 
This structure implements an update policy of using
|Dom.div| operation without restrictions -- which is possible only if
the domain has such an unrestricted operation. A domain such as the integer
domain may still provide |Dom.div| of the same type, but that operation may
only be used when we are sure that the division is exact. Our type
sharing constraint expresses such domain-specific knowledge:
instantiating |DivisionUpdate| with |IntegerDomain| leads to a
compile-time error, when compiling the \emph{generator} code. Thus, in
some cases we can use module types for ``semantic'' constraints
that cannot normally be expressed via the types of module members.
\vspace*{-5pt}
\begin{code}
module GenIV5 = Gen(IntegerDomain)
   (GenericVectorContainer)(FullPivot)
   (FractionFreeUpdate(IntegerDomain)(GenericVectorContainer)(IDet))
   (OutDetRank(IntegerDomain)(GenericVectorContainer)(IDet)(Rank))
module GenFA1 = Gen(FloatDomain)
   (GenericArrayContainer)(RowPivot)
   (DivisionUpdate(FloatDomain)(GenericArrayContainer)(NoDet(FloatDomain)))
   (OutJustMatrix(FloatDomain)(GenericArrayContainer)(NoDet(FloatDomain)))
\end{code}
\vspace*{-5pt}
We can instantiate the |Gen| functor as shown above and inspect the generated
code, e.g., by printing |GenFA1.gen|. The code can then be ``compiled'' as 
|!. GenFA1.gen| or with off-shoring. The code for |GenIV5| (Appendix A) shows
full pivoting, determinant and rank tracking. The code for all these aspects is
fully inlined; no extra functions are invoked and no tests other than those
needed by the LU algorithm itself are performed. The GE?? function returns a
triple |int array * int * int| of the U-factor, determinant and the rank. The
code generated by |GenFA1| (Appendix B) shows absolutely no traces of
determinant tracking: no declaration of spurious variables, no extra tests,
etc. The code appears as if the determinant tracking aspect did not exist
at all. The generated code for the above and other instantiations of
|Gen| can be examined at \cite{metamonadsURL}. The website also 
contains benchmark code and timing comparisons.

@@

\bibliography{metamonads}
\bibliographystyle{elsart-num}
\section{Appendix 0}
\label{app:perform}
\oleg{grammar of our perform monad. Thank Lydia}

@@

# val resIV5 : ('a,
   Funct4.GenIV5.Ctr.contr ->
   Funct4.OutDetRank(Funct4.IntegerDomain)(Funct4.GenericVectorContainer)
                    (Funct4.IDet)(Funct4.Rank).res) code =
  .<fun a_405 ->
   let t_406 = (ref 0) in let t_407 = (ref 0) in
   let t_408 = {arr = (Array.copy a_405.arr)} (a_405) in
   let t_409 = a_405.m in let t_410 = a_405.n in
   let t_411 = (ref 1) in let t_412 = (ref 1) in
   while (((! t_407) < t_409) && ((! t_406) < t_410)) do
    let t_413 = (! t_406) in let t_414 = (! t_407) in
    let t_415 = (ref (None)) in
    let t_435 =

@@

      for j_431 = t_413 to (t_410 - 1) do
       for j_432 = t_414 to (t_409 - 1) do
        let t_433 = (t_408.arr).((j_431 * t_408.m) + j_432) in
        if (not (t_433 = 0)) then
         (match (! t_415) with
          | Some (i_434) ->
             if ((abs (snd i_434)) > (abs t_433)) then
              (t_415 := (Some ((j_431, j_432), t_433))) else ()
          | None -> (t_415 := (Some ((j_431, j_432), t_433))))

@@

      (match (! t_415) with
       | Some (i_416) ->  (* swapping of columns *)
          if ((snd (fst i_416)) <> t_414) then begin
           let a_424 = t_408.arr and nm_425 = (t_408.n * t_408.m)
           and m_426 = t_408.m in
           let rec loop_427 =
            fun i1_428 -> fun i2_429 ->
              if (i2_429 < nm_425) then
               let t_430 = a_424.(i1_428) in
               a_424.(i1_428) <- a_424.(i2_429);
               a_424.(i2_429) <- t_430;
               (loop_427 (i1_428 + m_426) (i2_429 + m_426))

@@

           (loop_427 t_414 (snd (fst i_416)));
           (t_412 := (~- (! t_412))) (* adjust the sign of det *)
          end else (); (* swapping of rows elided *)
          (Some (snd i_416))

@@

    (match t_435 with
     | Some (i_436) ->

@@

         for j_437 = (t_413 + 1) to (t_410 - 1) do
          if (not ((t_408.arr).((j_437 * t_408.m) + t_414) = 0)) then begin
           for j_438 = (t_414 + 1) to (t_409 - 1) do
            (t_408.arr).((j_437 * t_408.m) + j_438) <-
             ((((t_408.arr).((j_437 * t_408.m) + j_438) * (* elided *)

@@

           (t_408.arr).((j_437 * t_408.m) + t_414) <- 0

@@

         done; (t_411 := i_436)

@@

        (t_406 := ((! t_406) + 1)) (* advance the rank *)
     | None -> (t_412 := 0));
    (t_407 := ((! t_407) + 1))

@@

   (t_408, 
    if ((! t_412) = 0) then 0 (* adjust the sign of the determinant *)
    else if ((! t_412) = 1) then (! t_411)
    else (~- (! t_411)), (! t_406))>.

@@

# val resFA1 : ('a,
   Funct4.GenFA1.Ctr.contr ->
   Funct4.OutJustMatrix(Funct4.FloatDomain)(Funct4.GenericArrayContainer)
                       (Funct4.NoDet(Funct4.FloatDomain)).res) code =

@@

   let t_2 = (ref 0) in let t_3 = (ref 0) in

@@

    let t_8 = (! t_2) in let t_9 = (! t_3) in
    let t_10 = (ref (None)) in
    let t_16 =

@@

      for j_13 = t_8 to (t_7 - 1) do
       let t_14 = (t_5.(j_13)).(t_9) in
       if (not (t_14 = 0.)) then
        (match (! t_10) with
         | Some (i_15) ->
            if ((abs_float (snd i_15)) < (abs_float t_14)) then
             (t_10 := (Some (j_13, t_14)))

@@

         | None -> (t_10 := (Some (j_13, t_14))))

@@

      (match (! t_10) with
       | Some (i_11) -> (* swapping of rows *)
          if ((fst i_11) <> t_8) then begin
           let t_12 = t_5.(t_8) in
           t_5.(t_8) <- t_5.(fst i_11);
           t_5.(fst i_11) <- t_12; () end else ();
          (Some (snd i_11))

@@

    (match t_16 with
        begin (* elimination loop, elided *) end;

@@

>>>>

The initialization expressions |OD.Det.fin ()| and |IF.R.fin ()| are
the pre-flight tests, to be explained in Section XXX.

@@

As is apparent from the output choices, several different quantities
\emph{may} need to be tracked in a particular LU implementation.  We
therefore need to be able to conditionally generate variables
representing the tracking state, and weave in corresponding tracking
code. We may need to (independently) keep track of the rank, the
determinant and the permutation list.  The tracking state variables
then become part of the \emph{state} that is tracked by our monad.  To
have all this choice when needed, and yet have our code be modular and
composable as well as ensuring that the generated code does not
contain any abstraction artifacts, it is important to make this state
modular.  For example,
\begin{code}
module type DETERMINANT = sig
  type tdet = C.Dom.v ref
  type 'a lstate
  type ('pc,'v) lm = ('pc,'v) cmonad
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  type ('pc,'v) om = ('pc,'v) omonad
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  val decl : unit -> ('b,unit) lm (* could be unit rather than unit code...*)
  val upd_sign  : unit -> ('b,unit) om
  val zero_sign : unit -> ('b,unit) lm
  val acc       : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
  val get       : unit -> ('b,tdet) lm
  val set       : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
  val fin       : unit -> ('b,C.Dom.v) lm
end
\end{code}
\noindent  to track determinant we should be able to generate code
for: defining variables used for tracking (|decl|),
updating the sign or the absolute
value of the determinant, converting the tracking state
to the final determinant value of the type |outdet|. LU of a
floating-point matrix with no determinant tracking uses the
instantiation of |DETERMINANT| where |outdet| is |unit| and all the
functions of that module generate no code. For integer matrices, we
have to track some aspects of the determinant, even if we don't output
it. The determinant tracking aspect is complex because tracking
variables, if any, are to be declared at the beginning of LU; the sign
of the determinant has to be updated on each row or column
permutation; the value of the determinant should be updated per each
pivoting. We use |lstate| to pass the tracking state, e.g., a piece of
code for the value of the type |Dom.v ref|, among
various determinant-tracking functions. The |lstate| is a part of the
overall monadic state. Other aspects, e.g., rank tracking, may use the
monadic state for passing of rank tracking variables. To be able to
compose determinant and rank tracking functors -- each of which may
(or may not) use the monadic state for passing its own data -- we make
extensive use of open records (a list of polymorphic variants
appeared to be the easiest way to implement such a union, in a purely
functional way). This lets us freely compose determinant-tracking,
rank-tracking, and other aspects.

We have several instances of Det
\begin{code}
module NoDet =
  struct
  type tdet = C.Dom.v ref
  type 'a lstate = unit
  let decl () = unitL
  let upd_sign () = ret None
  let zero_sign () = unitL
  let acc _ = unitL
  let get () = ret (liftRef C.Dom.zeroL) (* hack alert! *)
  let set _ = unitL
  let fin () = failwith "Determinant is needed but not computed"
  type ('pc,'v) lm = ('pc,'v) cmonad
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  type ('pc,'v) om = ('pc,'v) omonad
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
end

module AbstractDet =
  struct
  open C.Dom
  type tdet = v ref
  (* the first part of the state is an integer: which is +1, 0, -1:
     the sign of the determinant *)
  type 'a lstate = ('a,int ref) abstract * ('a,tdet) abstract
  let fin = fun () -> perform
      (det_sign,det) <-- mo_lookup ip;
      ifM (Logic.equalL (liftGet det_sign) Idx.zero) (ret zeroL)
      (ifM (Logic.equalL (liftGet det_sign) Idx.one) (ret (liftGet det))
          (ret (uminusL (liftGet det))))
end
\end{code}

@@

\begin{code2}
module GenGE(F : FEATURES) = struct
    module O = F.Output(F)

@@

   let gen input = perform
          (mat, r, c, rmar) <-- init input;
          seqM 
            (forward_elim (mat, r, c, rmar))
            (O.make_result mat)
end
\end{code2}

@@

In addition to the ``regular'' type sharing constraints shown in the
|Gen| functor, there are also ``semantic'' sharing constraints, shown
in the following structure of the |UPDATE| signature:
\vspace*{-5pt}\begin{code}
module DivisionUpdate
  (Dom:DOMAIN with type kind = domain_is_field)
  (C:CONTAINER2D)
  (Det:DETERMINANT with type indet=Dom.v) = struct ... end
\end{code}
\vspace*{-4pt} 
This structure implements an update policy of using
|Dom.div| operation without restrictions -- which is possible only if
the domain has such an unrestricted operation. A domain such as the integer
domain may still provide |Dom.div| of the same type, but that operation may
only be used when we are sure that the division is exact. Our type
sharing constraint expresses such domain-specific knowledge:
instantiating |DivisionUpdate| with |IntegerDomain| leads to a
compile-time error, when compiling the \emph{generator} code. Thus, in
some cases we can use module types for ``semantic'' constraints
that cannot normally be expressed via the types of module members.
\vspace*{-5pt}
\begin{code}
module GVC_I = GenericVectorContainer(IntegerDomainL)
module G_GVC_I = GenLA(GVC_I)
open G_GVC_I
open G_GVC_I.GE
module GenIV5 = GenGE(struct 
    module Det = AbstractDet
    module PivotF = FullPivot
    module PivotRep = PermList
    module Update = FractionFreeUpdate
    module Input = InpJustMatrix
    module Output = OutDetRank end)

module GAC_F = GenericArrayContainer(FloatDomainL)
module G_GAC_F = GenLA(GAC_F)
open G_GAC_F
open G_GAC_F.GE

module GenFA1 = GenGE(struct
    module Det = NoDet
    module PivotF = RowPivot
    module PivotRep = PermList
    module Update = DivisionUpdate
    module Input = InpJustMatrix 
    module Output = OutJustMatrix end)
module GenFA9 = GenGE(struct 
    module Det = NoDet
    module PivotF = RowPivot
    module PivotRep = PermList
    module Update = DivisionUpdate
    module Input = InpJustMatrix
    module Output = Out_LU_Packed end)

let resIV5 = instantiate GenIV5.gen ;;
let rIV5 = runit {pf =  resIV5 };;

\end{code}
|GenFA1| is no-frills GE; GenFA9 has LU part

We can instantiate the |Gen| functor as shown above and inspect the generated
code, e.g., by printing |GenFA1.gen|. The code can then be ``compiled'' as 
|!. GenFA1.gen| or with off-shoring. The code for |GenIV5| (Appendix A) shows
full pivoting, determinant and rank tracking. The code for all these aspects is
fully inlined; no extra functions are invoked and no tests other than those
needed by the LU algorithm itself are performed. The GE?? function returns a
triple |int array * int * int| of the U-factor, determinant and the rank. The
code generated by |GenFA1| (Appendix B) shows absolutely no traces of
determinant tracking: no declaration of spurious variables, no extra tests,
etc. The code appears as if the determinant tracking aspect did not exist
at all. The generated code for the above and other instantiations of
|Gen| can be examined at \cite{metamonadsURL}. The website also 
contains benchmark code and timing comparisons.

@@

\bibliography{metamonads}
\bibliographystyle{elsart-num}
\section{Appendix 0}
\label{app:perform}
Grammar of our perform monad
We support four different constructs to introduce a monadic
expressions.

\begin{code}
  perform exp
  perform exp1; exp2
  perform x <-- exp1; exp2
  perform let x = foo in exp
\end{code}

which is almost literally the grammar of the Haskell's "do"-notation,
with the differences that Haskell uses |do| and |<-| where we use
|perform| and |<--|.

We support not only |let x = foo in ...|  expressions but arbitrarily
complex let-expressions, including |let rec| and |let module|.

The actual bind function of the monad defaults to |bind| and the
match-failure function to |failwith| (only used for refutable
patterns; see below).  To select a different function, use the
extended forms of |perform|. For example, use the function named 
|bind| from module |Mod|.  In
addition use the module's |failwith|-function in refutable patterns.
\begin{code}
        perform with module Mod in exp2
\end{code}
The code has full explanation...

@@

val resIV5 : ('a, GVC_I.contr -> GenIV5.O.res) code =
   let t_2 = (ref 0) in
   let t_3 = (ref 0) in
   let t_4 = {arr = (Array.copy a_1.arr)} (a_1) in
   let t_5 = a_1.m in
   let t_6 = a_1.n in
   let t_7 = (ref 1) in
   let t_8 = (ref 1) in
   while (((! t_3) < t_5) && ((! t_2) < t_6)) do
    let t_13 = (! t_2) in
    let t_14 = (! t_3) in
    let t_15 = (ref (None)) in
    let t_34 =

@@

      for j_30 = t_13 to (t_6 - 1) do
       for j_31 = t_14 to (t_5 - 1) do
        let t_32 = (t_4.arr).((j_30 * t_4.m) + j_31) in
        if (t_32 <> 0) then
         (match (! t_15) with
          | Some (i_33) ->
             if ((abs (snd i_33)) > (abs t_32)) then
              (t_15 := (Some ((j_30, j_31), t_32)))
          | None -> (t_15 := (Some ((j_30, j_31), t_32))))

@@

      (match (! t_15) with
       | Some (i_16) -> (* swapping of columns *)
          if ((snd (fst i_16)) <> t_14) then begin
           let a_23 = t_4.arr
           and nm_24 = (t_4.n * t_4.m)
           and m_25 = t_4.m in
           let rec loop_26 =
            fun i1_27 ->
             fun i2_28 ->
              if (i2_28 < nm_24) then
               let t_29 = a_23.(i1_27) in
               a_23.(i1_27) <- a_23.(i2_28);
               a_23.(i2_28) <- t_29;
               (loop_26 (i1_27 + m_25) (i2_28 + m_25))

@@

           (loop_26 t_14 (snd (fst i_16)));
           (t_8 := (~- (! t_8))) (* adjust the sign of det *)
          end else ();
          if ((fst (fst i_16)) <> t_13) then begin (* swapping of rows?*)
           let a_17 = t_4.arr
           and m_18 = t_4.m in
           let i1_19 = (t_13 * m_18)
           and i2_20 = ((snd (fst i_16)) * m_18) in
           for i_21 = 0 to (m_18 - 1) do
            let t_22 = a_17.(i1_19 + i_21) in
            a_17.(i1_19 + i_21) <- a_17.(i2_20 + i_21);
            a_17.(i2_20 + i_21) <- t_22
           (t_8 := (~- (! t_8)))
          end else ();
          (Some (snd i_16))

@@

    (match t_34 with
     | Some (i_35) ->

@@

         for j_36 = (t_13 + 1) to (t_6 - 1) do
          let t_37 = (t_4.arr).((j_36 * t_4.m) + t_14) in
          if (t_37 <> 0) then begin
           for j_38 = (t_14 + 1) to (t_5 - 1) do
            (t_4.arr).((j_36 * t_4.m) + j_38) <-
             ((((t_4.arr).((j_36 * t_4.m) + j_38) * i_35) -
                ((t_4.arr).((t_13 * t_4.m) + j_38) * t_37)) / (! t_7))

@@

           (t_4.arr).((j_36 * t_4.m) + t_14) <- 0

@@

         (t_7 := i_35)

@@

        (t_2 := ((! t_2) + 1)) (* advance the rank *)
     | None -> (t_8 := 0));

@@

   (t_4,
    if ((! t_8) = 0) then 0 (* adjust the sign of the determinant *)
    else if ((! t_8) = 1) then (! t_7)
    else (~- (! t_7)), (! t_2))>.

@@

val resFA1 : ('a, GAC_F.contr -> GenFA1.O.res) code =

@@

   let t_2 = (ref 0) in
   let t_3 = (ref 0) in

@@

    let t_10 = (! t_2) in
    let t_11 = (! t_3) in
    let t_12 = (ref (None)) in
    let t_18 =

@@

      for j_15 = t_10 to (t_7 - 1) do
       let t_16 = (t_5.(j_15)).(t_11) in
       if (t_16 <> 0.) then
        (match (! t_12) with
            if ((abs_float (snd i_17)) < (abs_float t_16)) then
             (t_12 := (Some (j_15, t_16)))

@@

         | None -> (t_12 := (Some (j_15, t_16))))

@@

      (match (! t_12) with (* swapping of rows *)
       | Some (i_13) ->
          if ((fst i_13) <> t_10) then
           let t_14 = t_5.(t_10) in
           t_5.(t_10) <- t_5.(fst i_13);
           t_5.(fst i_13) <- t_14
          else ();
          (Some (snd i_13))

@@

    (match t_18 with
     | Some (i_19) -> (* elimination loop, elided?? *)
        begin
         for j_20 = (t_10 + 1) to (t_7 - 1) do
          let t_21 = (t_5.(j_20)).(t_11) in
          if (t_21 <> 0.) then begin
           for j_22 = (t_11 + 1) to (t_6 - 1) do
            (t_5.(j_20)).(j_22) <-
             ((t_5.(j_20)).(j_22) -. ((t_21 /. i_19) *. (t_5.(t_10)).(j_22)))
           (t_5.(j_20)).(t_11) <- 0.
         ()

@@

\begin{code2}
val resFA9 : ('a, GAC_F.contr -> GenFA9.O.res) code =
  .<fun a_1 ->
   let t_2 = (ref 0) in
   let t_3 = (ref 0) in
   let t_5 = (Array.map (fun x_4 -> (Array.copy x_4)) (Array.copy a_1)) in
   let t_6 = (Array.length a_1.(0)) in
   let t_7 = (Array.length a_1) in
   let t_8 = (ref ([])) in
   while (((! t_3) < t_6) && ((! t_2) < t_7)) do
    let t_9 = (! t_2) in
    let t_10 = (! t_3) in
    let t_11 = (ref (None)) in
    let t_17 =
     begin
      for j_14 = t_9 to (t_7 - 1) do
       let t_15 = (t_5.(j_14)).(t_10) in
       if (t_15 <> 0.) then
        (match (! t_11) with
         | Some (i_16) ->
            if ((abs_float (snd i_16)) < (abs_float t_15)) then
             (t_11 := (Some (j_14, t_15)))
            else ()
         | None -> (t_11 := (Some (j_14, t_15))))
       else ()
      done;
      (match (! t_11) with
       | Some (i_12) ->
          if ((fst i_12) <> t_9) then begin
           let t_13 = t_5.(t_9) in
           t_5.(t_9) <- t_5.(fst i_12);
           t_5.(fst i_12) <- t_13;
           (t_8 := ((RowSwap ((fst i_12), t_9)) :: (! t_8)))
          end else ();
          (Some (snd i_12))
       | None -> (None))
     end in
    (match t_17 with
     | Some (i_18) ->
        begin
         for j_19 = (t_9 + 1) to (t_7 - 1) do
          let t_20 = (t_5.(j_19)).(t_10) in
          if (t_20 <> 0.) then
           for j_21 = (t_10 + 1) to (t_6 - 1) do
            (t_5.(j_19)).(j_21) <-
             ((t_5.(j_19)).(j_21) -. ((t_20 /. i_18) *. (t_5.(t_9)).(j_21)))
           done
          else ()
         done;
         ()
        end;
        (t_2 := ((! t_2) + 1))
     | None -> ());
    (t_3 := ((! t_3) + 1))
   done;
   (t_5, (! t_8))>.
\end{code2}
